---
title: "5: Data Exploration"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2020"
output: pdf_document
geometry: margin=2.54cm
fig_width: 5
fig_height: 2.5
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Import and explore datasets in R
2. Graphically explore datasets in R
3. Apply data exploration skills to a real-world example dataset

## Opening discussion: why do we explore our data?

Why is data exploration our first step in analyzing a dataset? What information do we gain? How does data exploration aid in our decision-making for data analysis steps further down the pipeline?

Helps you find out if you even have the data you need to answer particular questions, or if you have too much data, what you might want to remove.

## Import data and view summaries

```{r, message = FALSE}
# 1. Set up your working directory
getwd()

# 2. Load packges
library(tidyverse)

# 3. Import datasets
USGS.flow.data <- read.csv("./Data/Processed/USGS_Site02085000_Flow_Processed.csv")

#View(USGS.flow.data)
# Alternate option: click on data frame in Environment tab

colnames(USGS.flow.data)
str(USGS.flow.data)
dim(USGS.flow.data)

# Check our date column
class(USGS.flow.data$datetime)
#Oh no, it isn;t actually an R

USGS.flow.data$datetime <- as.Date(USGS.flow.data$datetime, format = "%Y-%m-%d") 
#retelling R that our date column is a date, and the format that the data are in
#R has a lot of trouble with dates, make sure there are no NA values
class(USGS.flow.data$datetime)

```


## Visualization for Data Exploration

Although the `summary()` function is helpful in getting an idea of the spread of values in a numeric dataset, it can be useful to create visual representations of the data to help form hypotheses and direct downstream data analysis. Below is a summary of the useful types of graphs for data exploration.

Note: each of these approaches utilize the package "ggplot2". We will be covering the syntax of ggplot in a later lesson, but for now you should familiarize yourself with the functionality of what each command is doing.

### Bar Chart (function: geom_bar)

Visualize count data for categorical variables. 

```{r, fig.height = 3, fig.width = 4}
ggplot(USGS.flow.data, aes(x = discharge.mean.approval)) +
  geom_bar()
#making data plot with USGS flow data, specifying aesthetics
#Shows 4 x axis categories of approval rate
```

### Histogram (function: geom_histogram)

Visualize distributions of values for continuous numerical variables. What is happening in each line of code? Insert a comment above each line.

```{r, fig.height = 3, fig.width = 4}
#Lots of open space, highlighy skewed
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean))

#specified width of bins to take a closer look into discharge means
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean), binwidth = 10)

#Specifies how many bins there should be, become wider, and counts in each bin increased because more points included in each bin
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean), bins = 20)

#Specifies binwidth, and the scale of discharge means, may not be capturing all data, but shows most useful data exploration
ggplot(USGS.flow.data, aes(x = discharge.mean)) +
  geom_histogram(binwidth = 10) + 
  scale_x_continuous(limits = c(0, 500))
  
#Now making one for gauge height
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = gage.height.mean))

#Naming plot saves the figure into the environment
gageplot<-ggplot(USGS.flow.data) +
  geom_histogram(aes(x = gage.height.mean))
#printing it allows it to show up and will be in knitted document
print(gageplot)
```
### Frequency line graph (function: geom_freqpoly)

An alternate to a histogram is a frequency polygon graph (distributions of values for continuous numerical variables). Instead of displaying bars,  counts of continuous variables are displayed as lines. This is advantageous if you want to display multiple variables or categories of variables at once.

```{r, fig.height = 3, fig.width = 4}
#Creates one plot with distributions of various things, specifying them as different color or line type so you can distinguish them
#0,10 specifies x axis
#lty 2 is a dashed line
ggplot(USGS.flow.data) +
  geom_freqpoly(aes(x = gage.height.mean), bins = 50) +
  geom_freqpoly(aes(x = gage.height.min), bins = 50, color = "lightsteelblue4") +
  geom_freqpoly(aes(x = gage.height.max), bins = 50,  lty = 2) +
  scale_x_continuous(limits = c(0, 10))

summary(USGS.flow.data$gage.height.max)
summary(USGS.flow.data$gage.height.min)
summary(USGS.flow.data$gage.height.mean)
#Less NA's in mean, so data more frequently recording mean than min and max, might be more beneficial to only graph those that have all 3

#Setting color based on approval, and put legend at the top
ggplot(USGS.flow.data) +
  geom_freqpoly(aes(x = gage.height.mean, color = gage.height.mean.approval), bins = 50) +
  scale_x_continuous(limits = c(0, 10)) +
  theme(legend.position = "top")

```
### Box-and-whisker plots (function: geom_boxplot, geom_violin)

A box-and-whisker plot is yet another alternative to histograms (distributions of values for continuous numerical variables). These plots consist of: 

* A box from the 25th to the 75th percentile of the data, called the interquartile range (IQR).

* A bold line inside the box representing the median value of the data. Whether the median is in the center or off to one side of the IQR will give you an idea about the skewness of your data.

* A line outside of the box representing values falling within 1.5 times the IQR. 

* Points representing outliers, values that fall outside 1.5 times the IQR. 

An alternate option is a violin plot, which displays density distributions, somewhat like a hybrid of the box-and-whiskers and the frequency polygon plot.

```{r, fig.height = 3, fig.width = 4}
#Doesn't show large differences in sample sizes** shows factor levels of approval
ggplot(USGS.flow.data) +
  geom_boxplot(aes(x = gage.height.mean.approval, y = gage.height.mean))

#can express continuous data instead of factor data
ggplot(USGS.flow.data) +
  geom_boxplot(aes(x = gage.height.mean, y = discharge.mean, group = cut_width(gage.height.mean, 1)))

#Violin Plots can help show differences in amount of data, still distribution of data
ggplot(USGS.flow.data) +
  geom_violin(aes(x = gage.height.mean.approval, y = gage.height.mean), 
              draw_quantiles = c(0.25, 0.5, 0.75))

#scale = "count" now shows the width of violin based on count of data recordings
ggplot(USGS.flow.data) +
  geom_violin(aes(x = gage.height.mean.approval, y = gage.height.mean), 
              draw_quantiles = c(0.25, 0.5, 0.75) , 
              scale = "count")
```

### Scatterplot (function: geom_point)
Visualize relationships between continuous numerical variables.

```{r, fig.height = 3, fig.width = 4}
ggplot(USGS.flow.data) +
  geom_point(aes(x = discharge.mean, y = gage.height.mean))

ggplot(USGS.flow.data) +
  geom_point(aes(x = datetime, y = discharge.mean))
```

Question: under what circumstances would it be beneficial to use each of these graph types (bar plot, histogram, frequency polygon, box-and whisker, violin, scatterplot)?

> Answer: violin and box and whisker are great for distribution of data, scatter plot can show specfific distribution of data points, bar plot is great for counts, more general

## Ending discussion

What did you learn about the USGS discharge dataset today? What separate insights did the different graph types offer?
> Answer: 


How can multiple options for data exploration inform our understanding of our data?

> Answer: 


Do you see any patterns in the USGS data for the Eno River? What might be responsible for those patterns and/or relationships?

> Answer: 


