---
title: "8: Data Visualization Basics"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2020"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  #can specify where want outputs to be displayed
  chunk_output_type: console
---
Notes: Barplots are really only great for count data*

## Objectives
1. Perform simple data visualizations in the R package `ggplot`
2. Develop skills to adjust aesthetics and layers in graphs
3. Apply a decision tree framework for appropriate graphing methods

## Opening discussion
Effective data visualization depends on purposeful choices about graph types. The ideal graph type depends on the type of data and the message the visualizer desires to communicate. The best visualizations are clear and simple. My favorite resource for data visualization is [Data to Viz](https://www.data-to-viz.com/), which includes both a decision tree for visualization types and explanation pages for each type of data, including links to R resources to create them. Take a few minutes to explore this website.

## Set Up

```{r, message = FALSE, warning = FALSE}
getwd()
library(tidyverse)
install.packages("ggridges")
library(ggridges)

PeterPaul.chem.nutrients <- 
  read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
PeterPaul.chem.nutrients.gathered <-
  read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv")
EPAair <- read.csv("./Data/Processed/EPAair_O3_PM25_NC1819_Processed_KS.csv")

EPAair$Date <- as.Date(EPAair$Date, format = "%Y-%m-%d")
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")
class(EPAair$Date)
```

## ggplot

ggplot, called from the package `ggplot2`, is a graphing and image generation tool in R. This package is part of tidyverse. While base R has graphing capabilities, ggplot has the capacity for a wider range and more sophisticated options for graphing. 
ggplot has only a few rules: 

* The first line of ggplot code always starts with `ggplot()`
* A data frame must be specified within the `ggplot()` function. Additional datasets can be specified in subsequent layers.
* Aesthetics must be specified, most commonly x and y variables but including others. Aesthetics can be specified in the `ggplot()` function or in subsequent layers.
* Additional layers must be specified to fill the plot. 

### Geoms
Type of plot, which we need to specify
Here are some commonly used layers for plotting in ggplot: 

* geom_bar
* geom_histogram
* geom_freqpoly
* geom_boxplot
* geom_violin
* geom_dotplot
* geom_density_ridges
* geom_point
* geom_errorbar
* geom_smooth
* geom_line
* geom_area
* geom_abline (plus geom_hline and geom_vline)
* geom_text

### Aesthetics
Here are some commonly used aesthetic types that can be manipulated in ggplot: 

* color
* fill
* shape
* size
* transparency

### Plotting continuous variables over time: Scatterplot and Line Plot
```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
ggplot(EPAair, aes(x = Date, y = Ozone)) + 
  geom_point()
#ggplot(which dataframe, aesthetics of x axis and y axis)+geom_point
#geompoint specifies which kind of plot to make, use a plus symbol to do this, without specifying it will just graph the x and y axis that had been specified

O3plot <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone))
print(O3plot) #print allows it to show up and appear in pdf when knit this

# Fix this code
O3plot2 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone, color = "blue"))
print(O3plot2)
#Need to keep the color outside of the aes(x,y)
#this is correct instead
O3plot2 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone), color = "blue")
print(O3plot2)

# Add additional variables
#This color aesthetic telling it to make the site names a different color
#as.factor makes Year a categorical(factor instead of integer) variable- doesn't change it from an integer in actual data frame
PMplot <- 
  ggplot(EPAair, aes(x = Month, y = PM2.5, shape = as.factor(Year), 
                     color = Site.Name)) +
  geom_point()
print(PMplot)
#this is not aesthetically worthy


# Separate plot with facets
#not using color-sitename
#facet wrap separating by site name so each has its own graph
#for all facets, specifies all rows
PMplot.faceted <-
  ggplot(EPAair, aes(x = Month, y = PM2.5, shape = as.factor(Year))) +
  geom_point() +
  facet_wrap(vars(Site.Name), nrow = 3)
print(PMplot.faceted)

# Filter dataset within plot building and facet by multiple variables
#subsetting is "filtering" by site name in this case
#or  ggplot(subset(EPAair,Site.Name %in% c("Clemens Middle", "", "")),
PMplot.faceted2 <- 
  ggplot(subset(EPAair, Site.Name == "Clemmons Middle" | Site.Name == "Leggett" |
                  Site.Name == "Bryson City"), 
         aes(x = Month, y = PM2.5)) + 
  geom_point() +
  facet_grid(Site.Name ~ Year) 
print(PMplot.faceted2)

# Plot true time series with geom_line
#time series must be within one spatial construct
PMplot.line <- 
  ggplot(subset(EPAair, Site.Name == "Leggett"), 
         aes(x = Date, y = PM2.5)) +
  geom_line() +
  geom_point()
print(PMplot.line)

#If there are gaps in time in data, R will still connect
#NA's will supply an actual break in data and not connect them



# Exercise: build your own scatterplots of PeterPaul.chem.nutrients

# 1. 
surftemps<- ggplot(subset(PeterPaul.chem.nutrients, depth==0)) +
  geom_point(aes(x = daynum, y = temperature_C, color = year4)) +
               facet_grid(vars(lakename))
             print(surftemps)
# Plot surface temperatures by day of  year. 
#Subset the dataset by depth=0 for surface water
# Color your points by year, and facet by lake in two rows.

             
#2. 
tempdate<- ggplot(PeterPaul.chem.nutrients)+
      geom_point(aes(x= sampledate, y= temperature_C, color = depth), 
                 size = 0.5)
print(tempdate)
    # Plot temperature by date. Color your points by depth.
# Change the size of your point to 0.5
#not a true time series because it's at multiple depths and for two lakes, a true time series needs to be at one spot over time


```
### Plotting the relationship between two continuous variables: Scatterplot

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
lightvsDO <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point()
print(lightvsDO)

# Adjust axes
#Huge outliers that are not real, so can readjust axes to focus in on the real data that is of interest
lightvsDOfixed <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point() +
  xlim(0, 250) +
  ylim(0, 20)
print(lightvsDOfixed)
#High irradiance levels, DO doesn;t change much, but at really low irradiance, large variance in DO values
#low low light, variability in Oxygen


# Depth in the fields of limnology and oceanography is on a reverse scale
#helps show cross section of lake- surface depth on top to bottom
tempvsdepth <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth)) +
  #ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth, color = daynum)) +
  #can add in color aes as third variable to see changes of temp and depth over seasons
  geom_point() +
  scale_y_reverse()
print(tempvsdepth)
#showing warmer temps at higher depths

NvsP <-
  ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = tn_ug, color = depth)) +
  geom_point() +
  geom_smooth(method = lm) + 
  #geomsmooth gives line of best fit, specifiying a linear model
  #geo_smooth(method=lm, se=FALSE) will rid of confidence interval around model line
  geom_abline(aes(slope = 16, intercept = 0))
#geom_abline is showing a known slope/relationship, here N should be 16 to 0 P
print(NvsP)

# Exercise: Plot relationships between air quality measurements

# 1. 
# Plot AQI values for ozone by PM2.5, colored by latitude 
# Make the points 50 % transparent
# Add a line of best fit for the linear regression of these variables.

  
```

### Plotting continuous vs. categorical variables
A traditional way to display summary statistics of continuous variables is a bar plot with error bars. Let's explore why this might not be the most effective way to display this type of data. Navigate to the Caveats page on Data to Viz (https://www.data-to-viz.com/caveats.html) and find the page that explores barplots and error bars. 

What might be more effective ways to display the information? Navigate to the boxplots page in the Caveats section to explore further.

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Box and whiskers plot
Nutrientplot3 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_boxplot(aes(color = nutrient)) # Why didn't we use "fill"?
print(Nutrientplot3)

# Dot plot
Nutrientplot4 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_dotplot(aes(color = nutrient, fill = nutrient), binaxis = "y", binwidth = 1, 
               stackdir = "center", position = "dodge", dotsize = 2) #
print(Nutrientplot4)

# Violin plot
Nutrientplot5 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_violin(aes(color = nutrient)) #
print(Nutrientplot5)

# Frequency polygons
# Using a tidy dataset
Nutrientplot6 <-
  ggplot(PeterPaul.chem.nutrients) +
  geom_freqpoly(aes(x = tn_ug), color = "darkred") +
  geom_freqpoly(aes(x = tp_ug), color = "darkblue") +
  geom_freqpoly(aes(x = nh34), color = "blue") +
  geom_freqpoly(aes(x = no23), color = "royalblue") +
  geom_freqpoly(aes(x = po4), color = "red") 
print(Nutrientplot6)

# Using a gathered dataset
Nutrientplot7 <-   
  ggplot(PeterPaul.chem.nutrients.gathered) +
  geom_freqpoly(aes(x = concentration, color = nutrient))
print(Nutrientplot7)

# Frequency polygons have the risk of becoming spaghetti plots. 
# See https://www.data-to-viz.com/caveat/spaghetti.html for more info.

# Ridgeline plot
Nutrientplot6 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(y = nutrient, x = concentration)) +
  geom_density_ridges(aes(fill = lakename), alpha = 0.5) #alpha was the transparency piece
print(Nutrientplot6)

# Exercise: Plot distributions of AQI values for EPAair

# 1. 
# Create several types of plots depicting PM2.5, divided by year. 
# Choose which plot displays the data best and justify your choice. 


```
