---
title: "10: Generalized Linear Models (T-Test)"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2020"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Describe the components of the generalized linear model (GLM)
2. Apply special cases of the GLM (t-test) to real datasets
3. Interpret and report the results of t-tests in publication-style formats

## Set up
```{r, message = FALSE}
getwd()
library(tidyverse)

EPAair <- read.csv("./Data/Processed/EPAair_O3_PM25_NC1819_Processed_KS.csv")

# Set date to date format
EPAair$Date <- as.Date(EPAair$Date, format = "%Y-%m-%d")

# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

## Generalized Linear Models (GLMs)

The one-sample test (model of the mean), two-sample t-test, analysis of variance (ANOVA), and linear regression are all special cases of the **generalized linear model** (GLM). The GLM also includes analyses not covered in this class, including logistic regression, multinomial regression, chi square, and log-linear models. The common characteristic of general linear models is the expression of a continuous response variable as a linear combination of the effects of categorical or continuous explanatory variables, plus an error term that expresses the random error associated with the coefficients of all explanatory variables. The explanatory variables comprise the deterministic component of the model, and the error term comprises the stochastic component of the model. Historically, artificial distinctions were made between linear models that contained categorical and continuous explanatory variables, but this distinction is no longer made. The inclusion of these models within the umbrella of the GLM allows models to fit the main effects of both categorical and continuous explanatory variables as well as their interactions. 

### Choosing a model from your data: A "cheat sheet"

**T-test:** Continuous response, one categorical explanatory variable with two categories (or comparison to a single value if a one-sample test).

**One-way ANOVA (Analysis of Variance):** Continuous response, one categorical explanatory variable with more than two categories.

**Two-way ANOVA (Analysis of Variance)** Continuous response, two categorical explanatory variables.

**Single Linear Regression** Continuous response, one continuous explanatory variable.

**Multiple Linear Regression** Continuous response, two or more continuous explanatory variables.

**ANCOVA (Analysis of Covariance)** Continuous response, categorical explanatory variable(s) and  continuous explanatory variable(s).

If multiple explanatory variables are chosen, they may be analyzed with respect to their **main effects** on the model (i.e., their separate impacts on the variance explained) or with respsect to their **interaction effects,** the effect of interacting explanatory variables on the model. 

### Assumptions of the GLM

The GLM is based on the assumption that the data residuals approximate a normal distribution (or a linearly transformed normal distribution). We will discuss the non-parametric analogues to several of these tests if the assumptions of normality are violated. For tests that analyze categorical explanatory variables, the assumption is that the variance in the response variable is equal among groups. Note: environmental data often violate the assumptions of normality and equal variance, and we will often proceed with a GLM even if these assumptions are violated. In this situation, justifying the decision to proceed with a linear model must be made. 

## T-Test
### One-sample t-test
The object of a one sample test is to test the null hypothesis that the mean of the group is equal to a specific value. For example, we might ask ourselves (from the EPA air quality processed dataset): 

Are Ozone levels below the threshold for "good" AQI index (0-50)?

```{r}

summary(EPAair$Ozone)
EPAair.subsample <- sample_n(EPAair, 5000) 
#taking random subsample (5000) of those data points

# Evaluate assumption of normal distribution
shapiro.test((EPAair.subsample$Ozone))
#assumption that data is normally distributed, but here p value is really really small, so possibility of normal distribution is very low, "data is not well proximated by normal distribution"

ggplot(EPAair, aes(x = Ozone)) +
  geom_histogram() 
#long right skew

qqnorm(EPAair$Ozone); qqline(EPAair$Ozone)
#compairing what we have vs what it would be if normal distribution

O3.onesample <- t.test(EPAair$Ozone, mu = 50, alternative = "less")
#saving this as an object so can call it up again
#giving one ozone column, comparing to a mu of 50, alternative hypoth is that the mean is less than 50
#if don't specify less, R will assume the alternative is greater or less than
O3.onesample
#from results... 95% confidence interval that mean is negative infinity to 41.13
#mean of x
#pvalue is less than 0.05 it is significant so we reject the null hypothesis
#null hypothesis is that the mean is equal to 50


Ozone.plot <- ggplot(EPAair, aes(x = Ozone)) +
  #geom_density(stat = "count", fill = "gray") +  #this shows the count
  geom_density(fill = "gray") + #this shows density distribution
  geom_vline(xintercept = 50, color = "#238b45", lty = 2, size = 0.9) +
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0))
#this final line of code makes axes start at 0, sometimes R will make them start a little higher
print(Ozone.plot)
```

Write a sentence or two about the results of this test. Include both the results of the test and an interpretation that puts the findings in context of the resarch question.

> When determining the mean of the Ozone data, we hypothesize that the mean will be less than 50. Based on our t test, the pvalue is less than 0.05 and therefore significant, so we reject the null hypothesis (mean-50), and accept alternative hypothesis that the mean Ozone is less than 50. 

>Kateri's: EPA ozone measurements for 18-19 were significantly lower than 50, the AQI threshold for "good" air quality (one sample t-test; t=-57.98, p<0.0001). #Communicating what question actually was and exactly what they need to know to answer the question

### Two-sample t-test
The two-sample *t* test is used to test the hypothesis that the mean of two samples is equivalent. Unlike the one-sample tests, a two-sample test requires a second assumption that the variance of the two groups is equivalent. Are Ozone levels different between 2018 and 2019?

```{r}
shapiro.test(EPAair$Ozone[EPAair$Year == 2018]) #defining ozone measurements for when the year is X
shapiro.test(EPAair$Ozone[EPAair$Year == 2019])
var.test(EPAair$Ozone ~ EPAair$Year)
#variances are significantly different from each other

ggplot(EPAair, aes(x = Ozone, color = as.factor(Year))) +
  geom_freqpoly()

# Format as a t-test
O3.twosample <- t.test(EPAair$Ozone ~ EPAair$Year) #test Ozone by year
#alternative would be that means are different
#null that the means are equal

O3.twosample
#pvalue not significant, reject null that the means are equal
#provides a mean for 2018 and 2019

O3.twosample$p.value
#just call us pvalue of the test

# Format as a GLM (generalized linear model)
O3.twosample2 <- lm(EPAair$Ozone ~ EPAair$Year)
summary(O3.twosample2) #need to call it up using summary with a LM
#treating year as continuous response variable
#gives approx increase in ozone each year (0.8452)

plot(O3.twosample2)
#par(mfrow=c(2,2)) allows the 4 graphs to be plotted all in one, after use this will need to set it back to (1,1) to only graph one at a time
```

### Non-parametric equivalent of t-test: Wilcoxon test

When we wish to avoid the assumption of normality, we can apply *distribution-free*, or non-parametric, methods in the form of the Wilcoxon rank sum (Mann-Whitney) test. The Wilcoxon test replaces the data by their rank and calculates the sum of the ranks for each group. Notice that the output of the Wilcoxon test is more limited than its parametric equivalent.

```{r}
O3.onesample.wilcox <- wilcox.test(EPAair$Ozone, mu = 50, alternative = "less")
O3.onesample.wilcox
#less statistical robustness, simply gives p value and summary
O3.twosample.wilcox <- wilcox.test(EPAair$Ozone ~ EPAair$Year)
#alt hypoth- diff air qualities in different years
```

### Visualization and interpretation challenge

Create three plots, each with appropriately formatted axes and legends. Choose a non-default color palette.

1. geom_density of ozone divided by year (distinguish between years by adding transparency to the geom_density layer).
2. geom_boxplot of ozone divided by year . Add letters representing a significant difference between 2018 and 2019 (hint: stat_summary). 
3. geom_violin of ozone divided by year, with the 0.5 quantile marked as a horizontal line. Add letters representing a significant difference between 2018 and 2019. 

```{r}
1.
ozone1<-ggplot(EPAair,aes(x = Ozone, fill = as.factor(Year)))+
  geom_density(alpha=0.5)+
  scale_x_continuous(expand = c(0, 0)) + 
  scale_y_continuous(expand = c(0, 0))
print(ozone1)

2.
ozone2<- ggplot(EPAair,aes(x = as.factor(Year)))+
  geom_boxplot(aes(y=Ozone))
print(ozone2)

3.
ozone3<- ggplot(EPAair,aes(x = as.factor(Year)))+
  geom_violin(aes(y=Ozone))
print(ozone3)
```

Now, write a summary of your findings, incorporating statistical output, reference to the figure(s), and a contextual interpretation.

> 



