---
title: "Assignment 9: Spatial Analysis"
author: "Kathleen Mason"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on spatial analysis. 

## Directions
1. Use this document to create code for a map. You will **NOT** be turning in the knitted Rmd file this time, only the pdf output for a map.
2. When you have produced your output, submit **only** the pdf file for the map, without any code. Please name your file "StudentName_A09_Spatial.pdf".

The completed exercise is due on Thursday, March 19 at 1:00 pm.

## Create a map

You have three options for this assignment, and you will turn in just **one** final product. Feel free to choose the option that will be most beneficial to you. For all options, to earn full points you should use best practices for data visualization that we have covered in previous assignments (e.g., relabeling axes and legends, choosing non-default color palettes, etc.).

Here are your three options: 

1. Reproduce figure 1b from the spatial lesson, found in section 3.2.2. You may choose a state other than North Carolina, but your map should still contain the spatial features contained in figure 1b in the "img" folder. 

2. Create a new map that mixes spatial and tabular data, as in section 3.3 of the spatial lesson. You may use the maps created in the lesson as an example, but your map should contain data other than precipitation days per year. This map should include: 

* State boundary layer
* Basin boundary layer
* Gage layer
* Tabular data (as an aesthetic for one of the layers)

3. Create a map of any other spatial data. This could be data from the spatial lesson, data from our other course datasets (e.g., the Litter dataset includes latitude and longitude of trap sites), or another dataset of your choosing.  Your map should include: 

* One or more layers with polygon features (e.g., country boundaries, watersheds)
* One or more layers with point and/or line features (e.g., sampling sites, roads)
* Tabular data that correpond to one of the layers, specified as an aesthetic (e.g., total litter biomass at each trap, land cover class at each trap)

Hint: One package that may come in handy here is the `maps` package, which contains several options for basemaps that cover political and geologic boundaries.


#Loading Packages
```{r load.libraries, error=FALSE, message=FALSE, warning=FALSE}
 getwd()
library("readr")
library("dplyr")
library("tidyr")
library("ggplot2")
library("purrr")
# some spatial data packages
#install.packages("sf")
library("sf")
#install.packages("ggmap")
library("ggmap")
# some other packages
#install.packages("here")
library("here")
```

#Read in Data
```{r read.in.some.data}
getwd()
basins_nf_seplains_raw <- st_read(here("data", "spatial_data", "bas_nonref_SEPlains.shp"))

gages_raw <- st_read(here("data", "spatial_data", "gagesII_9322_sept30_2011.shp"))

southeast_state_bounds_raw <- st_read(here("data", "spatial_data", "southeast_state_bounds.shp"))

Hydrotabular_data_raw <- read.csv(here("data", "tabular_data", "conterm_hydro.txt"))
Climatetabular_data_raw <- read.csv(here("data", "tabular_data", "conterm_climate.txt"))
Geotabular_data_raw <- read.csv(here("data", "tabular_data", "conterm_geology.txt"))
Topotabular_data_raw <- read.csv(here("data", "tabular_data", "conterm_topo.txt"))
```


#Setting up Projections
```{r activity.setting.projecting}
# define proj4 string and EPSG code
my_proj4 <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"
my_epsg <- 5070

# setting projections
st_crs(basins_nf_seplains_raw)<-my_proj4
st_crs(basins_nf_seplains_raw)<-my_epsg

st_crs(gages_raw)<-my_proj4
st_crs(gages_raw)<-my_epsg

st_crs(southeast_state_bounds_raw)<-my_proj4
st_crs(southeast_state_bounds_raw)<- my_epsg

#Rename
basins<- basins_nf_seplains_raw

gages <- gages_raw

state_bounds <- southeast_state_bounds_raw

# check CRSs
st_crs(gages)
st_crs(basins)
st_crs(state_bounds)
```


#Convert to SF Object
```{r tabular.to.sf }
# read in tabular format
gages_tabular <- read_csv(here("data", "tabular_data", "gagesII_sept30_2011_tabular.csv")) %>%
  as.data.frame() # use this to drop the tibble classes (i.e., tbl_df and tbl)

class(gages_tabular)

# use names() to get spatial data columns
names(gages_tabular)
# LAT_GAGE is y value and LNG_GAGE is x value

# convert to sf object
gages_tabular_as_sf <- st_as_sf(gages_tabular, coords = c("LNG_GAGE", "LAT_GAGE"), crs = 4326, dim = "XY") # lat long so set CRS = WGS84

# check class and CRS
class(gages_tabular_as_sf)
head(gages_tabular_as_sf)
st_crs(gages_tabular_as_sf)
# we'll need to change that if we want to use it with state bounds and watershed boundaries

# check by plotting (more on plotting below)
pdf(here("outputs", "tabular_to_sf_check_today.pdf"), width = 11, height = 8.5)
ggplot() +
  geom_sf(data = gages_tabular_as_sf) +
  theme_bw()
dev.off()

```

#Dropping tabular data
```{r isolating.geometry}
southeast_state_bounds_geometry <- st_geometry(southeast_state_bounds)

head(southeast_state_bounds_geometry)
#dropped tabular data and now just a multi polygon

# notice the class is different now
class(southeast_state_bounds_geometry)
# it's an sfc class, specifically an sfc_MULTIPOLYGON class
```

#Drop geo??
```{r dropping.geometry}
# geometry data is attached
head(southeast_state_bounds)
southeast_state_bounds_tabular <- st_drop_geometry(southeast_state_bounds)

# geometry data is dropped
head(southeast_state_bounds_tabular)

# notice the class is different now
class(southeast_state_bounds_tabular)
# it's a data.frame class, we could convert to a tibble using as_tibble()
```

#Filtering through data we need
```{r activity.gages.select}
# select a state, here I'm choosing North Carolina (NC)
# add your code here
Georgia<- southeast_state_bounds %>% filter(NAME== "Georgia")%>%
  st_geometry() #keeps spatial data and drops attribute data, only has geometry data now
names(Georgia)
class(Georgia)
# check
# add your code here
head(Georgia)
# select watersheds that intersect with NC bounds
# add your code here

GA_basins_nf_seplains<-basins_nf_seplains%>%
  st_intersection(Georgia) #intersection same as clip in ARCGIS
# check
# add your code here
head(GA_basins_nf_seplains)

# select gages that fall within NC bounds
# add your code here
GA_gages<- gages %>%
  st_intersection(Georgia)



GAbuffer<-st_buffer(GA_gages, dist=100) #buffer around stream gages in VA

not_GA_gages<- st_difference(gages, GA) #stream gages not in VA
```



#Visualizing the data
```{r basic.sf.plotting}


pdf(here("outputs", "GA_test_fig.pdf"), width = 11, height = 8.5)
ggplot() +
  geom_sf(data = Georgia, fill = NA) +
  geom_sf(data = GA_basins_nf_seplains, fill= "Pink", alpha = 0.25) +
  geom_sf(data = GA_gages, lwd = 1) +
  theme_bw()
dev.off()

```


#Adding in Tabular Data
```{r tabular data input}

names(Hydrotabular_data_raw)


#joining hydro tabular and ga gages data via STAID
GA_gages_Topo <- GA_gages %>%
  left_join(Topotabular_data_raw, by = "STAID")
#One STAID is a factor and one is numeric, let's see which one is which
class(GA_gages$STAID) #factor
class(Topotabular_data_raw$STAID) #numeric
#Lets change Hydrotabular_data_raw STAID to a factor
Topotabular_data_raw$STAID<- as.factor(Topotabular_data_raw$STAID)
class(Topotabular_data_raw$STAID)
#It worked!!

#Now try to rejoin
GA_gages_Topo <- GA_gages %>%
  left_join(Topotabular_data_raw, by = "STAID")
#warning message: Column `STAID` joining factors with different levels, coercing to character vector 
#Let's see if it worked

#PROBLEM??
#***Ga_gages_Topo is putting NA's in all columns that were the topo data set




```

#Now we plot
```{r plotting with tabular and spatial}
pdf(here("outputs", "GA_gages_by_tabular_data.pdf"), width = 11, height = 8.5)
ggplot() +
  geom_sf(data = GA_gages_Topo, aes(color = SLOPE_PCT), size = 3) +
  scale_color_gradient(low = "white", high = "blue") +
  labs(color = "Percent of Stream kilometers Identified as Artificial") +
  geom_sf(data = Georgia, fill = NA) +
  theme_bw()
dev.off()


```